Summary Statistics

### 2025-03-03 extensive_tests/draft-sumstat-vignette.Rmd
Updating draft vignette summarystatistics

```{r, echo=FALSE}
## TODO: Update 
library(devtools)
load_all()
```

Interpretation of arguments:

 - tip can be interpreted as sample or replicate

 - structure can be interpreted as island or non-island

## Frequency of methylation states per genomic structure type

@@

\[
\overline{x}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}x_{i,t} \\
\overline{y}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}y_{j,t}
\]

, where $x_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $y_{j,t}$ is the frequency of state $p$ in non-island $j$ and tip $t$.

- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$ at tip $t$, respectively. We define the mean global frequency of $w$ and $z$ as:

\[
\overline{w}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}w_{i,t} \\
\overline{z}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}z_{j,t}
\]

, where $w_{i,t}$ is the frequency of state $p$ in island $i$ and tip $t$ and $z_{j,t}$ is the frequency of state $m$ in non-island $j$ and tip $t$.

These values can be computed with the functions `get_islandMeanFreqP`, `get_nonislandMeanFreqP`,`get_islandMeanFreqM` and `get_nonislandMeanFreqM` as in the following examples:

@@

### Mean SD of observed frequencies per structure type at tree tips

- **Partially-methylated sites**: Let $x_{i,t}$ and $y_{j,t}$ be the frequency of methylation state $p$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively. We define the standard deviation of the frequency of $x$ and $y$ at tip $t$ as:

$$
\sigma_{t}(x)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
\sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
$$

, where $\overline{x_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in islands and $\overline{y_{.,t}}$ is the mean frequency of state $p$ at tip $t$ in non-islands:

$$
\overline{x_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}x_{i,t} \\
\overline{y_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}x_{j,t}
$$

Therefore, the mean standard deviation across tips is:

$$
\hat{\sigma}(x)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(x) \\
\hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y) \\
$$

- **Methylated sites**: Let $w_{i,t}$ and $z_{j,t}$ be the frequency of methylation state $m$ in island $i \in\{1, \cdots, I\}$ and non-island $j \in\{1, \cdots, J\}$, respectively. We define the standard deviation of the frequency of $w$ and $z$ at tip $t$ as:

$$
\sigma_{t}(w)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
\sigma_{t}(z)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{i,t}-\overline{y_{.,t}})^2}
$$

, where $\overline{x_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in islands and $\overline{z_{.,t}}$ is the mean frequency of state $m$ at tip $t$ in non-islands:

$$
\overline{w_{.,t}}=\frac{1}{I}\sum_{i=1}^{I}w_{i,t} \\
\overline{z_{.,t}}=\frac{1}{J}\sum_{j=1}^{J}z_{j,t}
$$

Therefore, the mean standard deviation across tips is:

$$
\hat{\sigma}(w)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(w) \\
\hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z) \\
$$
These values can be computed with the functions `get_islandSDFreqP`, `get_nonislandSDFreqP`,`get_islandSDFreqM` and `get_nonislandSDFreqM` as in the following examples:

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5)) # tip 1
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

@@

# Mean correlation of a central segment of methylation states per genomic structure type

Let $s$ be the length of the 2 adjacent segments of central positions to consider.
Let $l$ represent a number of positions to exclude at the start and end of each island/non-island structure.
Define the index positions $p=l+1$ and $p' = s-l$.
Let $x_{i,t,[p,p'-1]}$ and $x_{i,t,[p+1,p']}$ be the methylation states $x \in \{u, p, m\}$ in island $i \in \{1, \cdots ,I\}$ at tip $t$ for the given segments. 
Let $y_{j,t,[p,p'-1]}$ and $y_{j,t,[p+1,p']}$ be the methylation states $y \in \{u, p, m\}$ in non-island $j \in \{1, \cdots ,J\}$ at tip $t$ for the given segments.

We define the mean correlation as:

\[
\overline{Cor(x)_{s,l}}=\frac1I\frac1T\sum_{i=1}^{I}\sum_{t=1}^{T}Cor(x_{i,t,[p:p'-1]}, x_{i,t,[p+1:p']}) \\
\overline{Cor(y)_{s,l}}=\frac1J\frac1T\sum_{j=1}^{J}\sum_{t=1}^{T}Cor(y_{j,t,[p:p'-1]}, y_{j,t,[p+1:p']})
\]

where correlations are computed only for segment pairs where the standard deviation of each segment is non-zero:

\[
\sigma(x_{i,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(x_{i,t,[p+1:p']})\neq 0 \\
\sigma(y_{j,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(y_{j,t,[p+1:p']})\neq 0
\]

@@

*Correlation measures how much the state of each site says about the next. So, when there is no variation (e.g. in middle segment state u of a site is followed by state u of the next), there is no correlation.*

@@

Let \(S_k\) be the set of sites in the genomic structure indexed by \(k\). For each site \(s \in S_k\), let the methylation state at a given tip \(t\) be denoted as \(m_{k,s,t}\), where \(m_{k,s,t} \in \{0, 0.5, 1\}\) represents the unmethylated, partially-methylated, or methylated state, respectively.

For a given cherry \((t_1, t_2)\), we compute the proportion of sites with differing methylation states between the two tips.

@@

Then, the proportion of sites with different methylation states in structure \(k\) for the cherry \((t_1, t_2)\) is given by:

@@

where \(|S_k|\) is the total number of sites in structure \(k\), and the sum counts the number of sites where the methylation state differs between tips \(t_1\) and \(t_2\).

@@

For a given cherry \((t_1, t_2)\), let \( \mathcal{I} \) be the set of indices for island structures and \( \mathcal{N} \) be the set of indices for non-island structures. Each structure \( k \) has a total number of sites \( |S_k| \), which we use as weights.

@@

- `MeanSiteFChange_cherry(tree, data, index_islands, index_nonislands)`. The function computes the per-cherry frequency of sites with different methylation states at each structure (islands and non-islands) using \code{get_siteFChange_cherry}. 
Then it calculates the weighted mean site frequency of methylation changes for each cherry separately for islands and non-islands.

```{r}
# Set example tree and methylation data
tree <- "((a:1.5,b:1.5):2,(c:2,d:2):1.5);"
data <- list(
    list(rep(1,10), rep(0,5), rep(1,8)), # tip a
    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip b
    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip c
    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # tip d 
# Set the index for islands and non-island structures
index_islands <- c(2)
index_nonislands <- c(1, 3)

@@

# Fitch

- `get_meanMeth_islands` computes the mean methylation state of all the islands. **Maybe this should allow excluding island borders**

>>>>

```{r, echo=FALSE}
## TODO: Update 
library(devtools)
load_all()
```
# Introduction

This document presents summary statistics related to the evolution of methylation 
states across different genomic structures. The statistics include mean 
frequencies, standard deviations, neighbor correlations of methylation states in 
island and non-island regions. Additionally, for analysis comparing different 
samples (comparison of tree tips), the included summary statistics compute
the mean per-site frequency of methylation changes and the Fitch estimate for
the minimum number of changes per islands leading to different global methylation
states.

The provided set of functions allows the computation of summary statistics for 
methylation data in a genomic region with structural categorization. In this 
vignette, we will refer to these structures as **islands** and 
**non-island structures**.

The `data` argument must be in one of the following formats, depending on the 
analysis:

1. **For analyses without tip comparison:** A list where each element represents 
the methylation states in a given structure (`data[[structure]]`):

```{r}
# Example: a single sample with 3 genomic structures
# (1) island with 10 partially-methylated sites
# (2) non-island with 5 methylated sites
# (3) island with 15 unmethylated sites

data <- list(rep(0.5, 10),  # Partially methylated
             rep(1,5),      # Methylated
             rep(0,15))     # Unmethylated
data
```

2. **For analyses comparing multiple tips in a phylogenetic tree:** A nested 
list where the first level represents the tips, and the second level represents 
the structures within each tip (`data[[tip]][[structure]]`).

```{r}
# Example: data from 3 tips of a tree,
# each with 3 genomic structures
data <- list(
    # Tip 1
    list(c(rep(0.5,5), rep(0,5)),  # 5 partially methylated, 5 unmethylated
         c(rep(1,4), 0.5),         # 4 methylated, 1 unmethylated
         c(rep(0,7), rep(0.5,8))), # 7 unmethylated, 8 partially methylated
    # Tip 2
    list(c(rep(0.5,9), 1), # 9 partially methylated, 1 methylated
         c(rep(0.5,4), 1), # 4 partially methylated, 1 methylated
         c(rep(0,8), rep(0.5,7))), # 8 unmethylated, 7 partially-methylated
    # Tip 3
    list(c(1, rep(0,8), 1), # first and last methylated, rest unmethylated
         c(rep(0.5,3), rep(1,2)), # 3 methylated, 1 unmethylated
         c(rep(0.5,15)))) # all partially methylated
```

In this case, `data` must be pre-filtered to include only sites present in all 
tips, ensuring a valid comparison between species or samples.

Regardless of the format, methylation values should be represented as follows:

- `0` for unmethylated sites,

- `0.5` for partially methylated sites,

- `1` for methylated sites.

Intermediate values (e.g., obtained from pooled empirical data) should be 
categorized before analysis.

Finally, when phylogenetic data is provided, the order of tips in `data` must 
match the tip order in the Newick tree format (left to right) or the order of 
`tree$tip.label()` when the tree is provided as a `phylo` object from the `ape` 
package.

```{r}
# Example tree in Newick format for the above data
newick_tree <- "((tip1:1, tip2:1):1, tip3:2);"

# Example tree as a phylo object from the ape package
library(ape)
phylo_tree <- read.tree(text = newick_tree)
phylo_tree$tip.label
```

# Summary Statistics

## Frequency of Methylation States per Genomic Structure Type

@@

$$
\overline{x}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}x_{i,t} \\
\overline{y}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}y_{j,t}
$$

where $x_{i,t}$ and $y_{j,t}$ represent the respective frequencies at each tip.

- **Methylated sites**: Similarly, let $w_{i,t}$ and $z_{j,t}$ be the frequency 
of methylation state $m$ in islands and non-islands. The mean global frequencies 
are:

$$
\overline{w}=\frac1I\sum_{i=1}^{I}\frac1T\sum_{t=1}^{T}w_{i,t} \\
\overline{z}=\frac1J\sum_{j=1}^{J}\frac1T\sum_{t=1}^{T}z_{j,t}
$$

These values can be computed using the following functions:

@@

### Mean SD of Observed Frequencies per Structure Type at Tree Tips

- **Partially-methylated sites**: We define the standard deviation of $x$ and 
$y$ at tip $t$ as:

$$
\sigma_{t}(x)=\sqrt{\frac{1}{I-1}\sum_{i=1}^{I}(x_{i,t}-\overline{x_{.,t}})^2} \\
\sigma_{t}(y)=\sqrt{\frac{1}{J-1}\sum_{j=1}^{J}(y_{j,t}-\overline{y_{.,t}})^2}
$$

where $\overline{x_{.,t}}$ and $\overline{y_{.,t}}$ represent mean frequencies 
at tip $t$. The mean standard deviation across tips is:

$$
\hat{\sigma}(x)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(x) \\
\hat{\sigma}(y)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(y)
$$

- **Methylated sites**: Similarly, for $w$ and $z$:

$$
\hat{\sigma}(w)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(w) \\
\hat{\sigma}(z)=\frac{1}{T}\sum_{t=1}^{T}\sigma_t(z)
$$

These values can be computed using:

```{r}
# 1 tip / sample / replicate
sample_n <- 1
index_islands <- c(1, 3)
index_nonislands <- c(2, 4)
data <- list(c(.5, .5, 0, 1, 1, .5), c(.5, 0, 1, .5), c(.5, .5, 0), c(0, 0, .5))
get_islandSDFreqP(index_islands, data, sample_n)
get_nonislandSDFreqP(index_nonislands, data, sample_n)
get_islandSDFreqM(index_islands, data, sample_n)
get_nonislandSDFreqM(index_nonislands, data, sample_n)
```

@@

## Mean Correlation of a Central Segment of Methylation States per Genomic Structure Type

Let $s$ be the length of two adjacent segments of central positions to consider. 
Let $l$ be the number of positions excluded at the start and end of each 
structure. Define:

$$
\overline{Cor(x)_{s,l}}=\frac1I\frac1T\sum_{i=1}^{I}\sum_{t=1}^{T}Cor(x_{i,t,[p:p'-1]}, x_{i,t,[p+1:p']}) \\
\overline{Cor(y)_{s,l}}=\frac1J\frac1T\sum_{j=1}^{J}\sum_{t=1}^{T}Cor(y_{j,t,[p:p'-1]}, y_{j,t,[p+1:p']})
$$

Correlations are computed only for segment pairs where standard deviation is 
non-zero:

$$
\sigma(x_{i,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(x_{i,t,[p+1:p']})\neq 0 \\
\sigma(y_{j,t,[p:p'-1]}) \neq 0 \text{ and } \sigma(y_{j,t,[p+1:p']})\neq 0
$$

@@

Note that correlation measures how much the state of each site says about the 
next. So, when there is no variation (e.g. in middle segment state u of a site 
is followed by state u of the next), there is no correlation.

@@

Let $S_k$ be the set of sites in the genomic structure indexed by $k$. 
For each site $s \in S_k$, let the methylation state at a given tip $t$ be 
denoted as $m_{k,s,t}$, where $m_{k,s,t} \in \{0, 0.5, 1\}$ represents the 
unmethylated, partially-methylated, or methylated state, respectively.

For a given cherry $(t_1, t_2)$, we compute the proportion of sites with 
differing methylation states between the two tips.

@@

Then, the proportion of sites with different methylation states in structure $k$ 
for the cherry $(t_1, t_2)$ is given by:

@@

where $|S_k|$ is the total number of sites in structure \(k\), and the sum 
counts the number of sites where the methylation state differs between tips 
$t_1$ and $t_2$.

@@

For a given cherry $(t_1, t_2)$, let $\mathcal{I}$ be the set of indices 
for island structures and $\mathcal{N}$ be the set of indices for non-island 
structures. Each structure $k$ has a total number of sites $|S_k|$,
which we use as weights.

@@

- `MeanSiteFChange_cherry(tree, data, index_islands, index_nonislands)`. The 
function computes the per-cherry frequency of sites with different methylation 
states at each structure (islands and non-islands) using `get_siteFChange_cherry`. 
Then it calculates the weighted mean site frequency of methylation changes for 
each cherry separately for islands and non-islands.

```{r}
# Set example tree and methylation data
tree <- "((a:1.5,b:1.5):2,(c:2,d:2):1.5);"
data <- list(
    list(rep(1,10), rep(0,5), rep(1,8)), # tip a
    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip b
    list(rep(1,10), rep(0.5,5), rep(0,8)), # tip c
    list(c(rep(0,5), rep(0.5, 5)), c(0, 0, 1, 1, 1), c(0.5, 1, rep(0, 6)))) # d

# Set the index for islands and non-island structures
index_islands <- c(2)
index_nonislands <- c(1, 3)

@@

### Fitch Parsimony Estimate for Minimum Global Methylation State Changes at CpG Islands

Let $I$ be the set of CpG islands in a given genomic region For each island 
$i \in I$, the mean methylation level across sites is computed with 
`get_meanMeth_islands(index_islands, data)` and categorized into one of three 
global methylation states—unmethylated ($u$), partially methylated ($p$), or 
methylated ($m$)—based on user-defined thresholds $u_{\text{thresh}}$ and 
$m_{\text{thresh}}$. This categorization is given by the function:

$$
\operatorname{categorize\_islandGlbSt}(\mu_i, u_{\text{thresh}}, m_{\text{thresh}}) =
\begin{cases}
  u, & \text{if } \mu_i \leq u_{\text{thresh}}, \\
  m, & \text{if } \mu_i \geq m_{\text{thresh}}, \\
  p, & \text{otherwise}.
\end{cases}
$$

Given a phylogenetic tree $T$ with tips corresponding to sampled species or 
individuals, we aim to estimate the minimum number of state changes required to 
explain the observed distribution of island methylation states across the tips. 
This is achieved using Fitch parsimony, a dynamic programming approach 
implemented in `compute_fitch(meth, tree)` with the following steps:

- Each tip $t$ is assigned the observed methylation state for each island $i$.
- Internal nodes recursively inherit possible state sets by computing the 
intersection of the state sets from their child nodes.
- If the intersection is empty (i.e., no common state between child nodes), the 
union of both state sets is assigned, and the number of changes is incremented.

The function `computeFitch_islandGlbSt(T, M)`, where $M$ is the matrix of 
categorized methylation states across tips, returns a vector $C$ where each entry 
$C_i$ represents the minimum number of changes required for island $i$ under the 
Fitch parsimony criterion.

```{r}
# Example with data from a single island structure 
# and three tips

tree <- "((bla:1,bah:1):2,booh:2);"

data <- list(
  #Tip 1
  list(c(rep(1,9), rep(0,1))), # m
  #Tip 2
  list(c(rep(0.5,10))), # p
  #Tip 3
  list(c(rep(0.5,9), rep(0.5,1)))) # p

index_islands <- c(1)

computeFitch_islandGlbSt(index_islands, data, tree, 
                         u_threshold = 0.1, m_threshold = 0.9)
```
```{r}
# Example: data from a genomic region consisting on 3 structures with 10 sites each
# one island, one non-island, one island
# and a tree with 8 tips
tree <- "(((a:1,b:1):1,(c:1,d:1):1):1,((e:1,f:1):1,(g:1,h:1):1):1);"
data <- list(
  #Tip 1
  list(c(rep(1,5), rep(0,5)), # p
       c(rep(0,9), 1), 
        c(rep(1,8), rep(0.5,2))), # m
  #Tip 2
  list(c(rep(0.5,9), rep(0.5,1)), # p
        c(rep(0.5,9), 1), 
       c(rep(0,8), rep(0.5,2))), # u
  #Tip 3
  list(c(rep(1,9), rep(0.5,1)), # m
       c(rep(0.5,9), 1), 
       c(rep(0.5,8), rep(0.5,2))), # p
  #Tip 4
  list(c(rep(1,9), rep(0.5,1)), # m
       c(rep(1,9), 0), 
       c(rep(0.5,8), rep(0.5,2))), # p
  #Tip 5
  list(c(rep(0,5), rep(0,5)), # u
       c(rep(0,9), 1), 
       c(rep(0.5,8), rep(0.5,2))), # p
  #Tip 6
  list(c(rep(0,9), rep(0.5,1)), # u
       c(rep(0.5,9), 1), 
       c(rep(1,8), rep(0.5,2))), # m
  #Tip 7
  list(c(rep(0,9), rep(0.5,1)), # u
       c(rep(0.5,9), 1), 
       c(rep(0,8), rep(0.5,2))), # u
  #Tip 8
  list(c(rep(0,9), rep(0.5,1)), # u
       c(rep(1,9), 0), 
       c(rep(0,9), rep(0.5,1)))) # u

index_islands <- c(1,3)

computeFitch_islandGlbSt(index_islands, data, tree, 
                         u_threshold = 0.1, m_threshold = 0.9)
```

For genomic regions containing multiple CpG islands, we can summarize the 
per-island estimates using the mean to obtain an overall measure of the minimum 
state changes.

```{r}
mean(computeFitch_islandGlbSt(index_islands, data, tree, 
                              u_threshold = 0.1, m_threshold = 0.9))
```
